---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r global settings, echo = FALSE}

knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
    
```

## Loading and preprocessing the data

We shall require **dplyr** and **ggplot2** libraries:
```{r load libraries}

library(dplyr)
library(ggplot2)

```

The following codes will load the data and convert the date column to Date type. Subsequently, a dplyr table data frame is created and we shall work from there:

```{r loading_data}

# load data
activity <- read.csv("activity.csv")

# convert the date column to date type
activity$date <- as.Date(activity$date)

# create a dplyr data frame to work with
actvt <- tbl_df(activity)

```

## What is mean total number of steps taken per day?

This code will aggregate the number of steps by day:

```{r total steps per day}

daily_steps <- actvt %>% group_by(date) %>% summarise(total = sum(steps))

```

This chart represents the histogram (not a bar plot!) of the total number of steps taken each day:

```{r histogram of daily total steps}

g <- ggplot(daily_steps, aes(total))
g + geom_histogram() + labs(x = "Total Steps per Day", y = "Frequency")

```

The mean of the steps per day is:

```{r the mean, echo = FALSE}
mean(daily_steps$total, na.rm = TRUE)
```

And the median for number of steps is:

```{r the median, echo = FALSE}
median(daily_steps$total, na.rm = TRUE)
```


## What is the average daily activity pattern?

This code will produce the aggregated average steps per interval across all days:

```{r total average steps per interval}

interval_steps <- actvt %>% group_by(interval) %>% summarise(avg = mean(steps, na.rm = TRUE))

interval_steps

```

This will make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r time series of interval average steps}

g <- ggplot(interval_steps, aes(interval, avg))
g + geom_line() + labs(x = "Interval", y = "Average Steps Across all Days")

```

From the graph we could see that the highest average is around the 750-900 minutes interval (to be exact at 835th interval)

## Imputing missing values

The following calculates the number of missing values in the dataset:

```{r missing values}

sum(is.na(activity$steps))

```

For the purpose of filling up the NA's, we shall use the average of that minutes interval. This shall create a new dataset with NAs being filled:


```{r fill NAs with interval average}

new_actvt <- actvt %>% group_by(interval) %>% mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))

```

This chart represents a new histogram of the total number of steps taken each day using the new data:

```{r histogram of daily total steps of a new dataset}

new_daily_steps <- new_actvt %>% group_by(date) %>% summarise(total = sum(steps))
g <- ggplot(new_daily_steps, aes(total))
g + geom_histogram() + labs(x = "Total Steps per Day", y = "Frequency")

```

The mean of the steps per day is:

```{r the new mean, echo = FALSE}
mean(new_daily_steps$total, na.rm = TRUE)
```

And the median for number of steps is:

```{r the new median, echo = FALSE}
median(new_daily_steps$total, na.rm = TRUE)
```

From the new mean and median, we could see that the impact is minimal and only on the median value.

## Are there differences in activity patterns between weekdays and weekends?

Adding a new column to indicate whether weekend or weekday:

```{r add weekdaytype column}

new_actvt <- new_actvt %>% mutate(weekdaytype = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

````

Plotting a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):

```{r weekday vs weekend plot}

new_interval_steps <- new_actvt %>% group_by(interval, weekdaytype) %>% summarise(avg = mean(steps, na.rm = TRUE))
g <- ggplot(new_interval_steps, aes(interval, avg))
g + geom_line() + labs(x = "Interval", y = "Average Steps Across all Days") + facet_grid(weekdaytype ~.)

```

From the graph it seems that activities do not differ much between weekdays and weekends.
